Using numerical modelling to investigate temperature and wave exposure as drivers

### 2019-11-08 literature_review.Rmd
added section to lit review, updated outputs, and added references to bib. file

---
title: A Transdiciplinary approach to developing a 3D spectral numerical coastal-flow model based on passive macroalgal rafting phenenomea to shed light on ambient and stochastic dispersal and connectivity of nearshore and offshore marine communities  around the coast of South Africa
author: "by RM Coppin"
output:
  word_document:
    fig_caption: yes
    toc: yes
    reference_docx: chapter_1_template.docx
  pdf_document:
    toc: no
bibliography: references.bib
---

@@

Marine ecosystems are maintained by a variety of complex interactions between abiotic and biotic variables such as temperature, wave exposure, pH, competition, and processes such as top-down and bottom-up control, predator-prey relationships and phenology (Doney et al. 2011; Harley et al. 2006; Burrows et al. 2011; McGowan et al. 1998). These abiotic and biotic variables, the interactions between them, and the various ecological processes, ultimately determine the community composition and ecological functioning of all ecosystems (Guimaraes & Coutinho 1996; Harley et al. 2006; Poloczanska et al. 2013; Jennings & Brander 2010; Polovina 2005; Johnson et al. 2011; Wernberg et al. 2016; Dayton et al. 1999). Climate directly and indirectly affects the way in which abiotic and biotic variables interact, but is often compounded by other impacts such as habitat destruction, pollution, and over-fishing (Blamey et al. 2012; Blamey et al. 2015). Temperature and wave exposure have been recognised as important variables with regards to climate-driven changes within the ocean (Guimaraes & Coutinho 1996; McGowan et al. 1998; McQuaid & Branch 1984; Laufkötter et al. 2015; Belkin 2009; Filbee-Dexter et al. 2018; Miller et al. 2011; Smale et al. 2011; Smale & Moore 2017). In order to persist and survive within variable and changing environments, organisms must either migrate, adapt, or die.

Seaweeds are sessile organisms which are unable to migrate to new areas when local environmental conditions become unsuitable and therefore are forced to adapt to new conditions in order to avoid expiration (Dayton 1985; Tegner et al. 1997). The main form of mortality for seaweeds is through mechanical dislodgement by wave action (Schiel et al. 2006; Seymour et al. 1989; Thomsen et al. 2004; Cavanaugh et al. 2011; Graham et al. 1997; Demes et al. 2013; Edwards & Estes 2006). Seaweeds, particularly brown seaweeds, are able to undergo rapid morphological adaptation to the hydrodynamic environment (Miller et al. 2011; Friedland & Denny 1995; Wing et al. 2007; Wernberg & Thomsen 2005; Fowler-Walker et al. 2006; Dudgeon & Johnson 1992; Blanchette et al. 2002). This allows seaweeds to reduce mortality through mechanical dislodgement by inducing morphology which reduces overall drag. Seaweeds that are unable to avoid mechanical dislodgement either raft out to sea or wash up onto beaches. Not all beach-cast kelp may have originated from a nearby kelp population and may have originated from other sites or regions of the coast through rafting and ocean currents (Malm et al. 2004; Emery & Tschudy 1941; Filbee-Dexter & Scheibling 2012; Hobday 2000). Therefore, because kelp morphology is specific to its local environment the morphological features may be able to indicate, within a certain amount of probability, what site or region it most likely originated from. In other words, beach-cast kelp may be used as proxy for investigating the flow of coastal currents. Using kelp as a proxy for determining its original location will be calibrated by means of a hydrodynamic model which will be designed from already existing SWAN and Delft3D models. This combined approach will allow investigation into flow regimes around the west and south-west coasts of South Africa and the role they play in subtidal ecology. Furthermore, the model could be used to determine transport of plastic pollution around the coast. For instance, micro-plastics are recognised as a threat to marine life (Seltenrich 2015; Azzarello & Van Vleet 1987; Wright et al. 2013; Derraik 2002; Andrady 2011; Fendall & Sewell 2009; Ivar Do Sul & Costa 2014; O ’donoghue & Marshall n.d.; Ballent et al. 2012), however very little is known about how microplastics may be transported along the coast.

@@

In the wave swept nearshore one would expect organisms to reflect the harsh hydrodynamic environment by being streamlined, small and amoured. This is certainly the case for an array of fauna which often comprise of hard, rigid bodies that are held firmly in place to the substratum such as limpets and isopods.

@@

```{r surface elevation figures, echo=FALSE, warning=FALSE, fig.cap="\\label{fig:surface elevation figures}The definition of a wave in a time record of the surface elevation with downward zero-crossings (upper panel) and upward zero-crossings (lower panel)."}

@@

```{r Hs_Tp_surface_elevation figure, echo=FALSE, warning=FALSE, fig.cap="The definition of wave height and wave period in a time record of the surface elevation (the wave is defined with downwrd zero-crossings."}

@@

# Hydrodynamic modelling
## Introduction

@@

# Calculating characteristics of drift
## Buoyancy of drifting objects in the oceans

@@

```{r BF_1, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 3, fig.height = 3, fig.cap='\\label{fig:BF_1}Conceptual model of bouyancy characteristics.'}

@@

The effects that wind produce on floating objects is sometimes referred to as "windage factors", and is difficult to accurately and empirically describe. The difficulty is due to accurate and measurements needed for the current velocity, wind velocity, wave height and wave direction. Furthermore, these factors do not act independently and instead influence one another which compounds the complexity of calculating wind effects on drifting objects. To overcome this challenge, measurements of current and wind velocity, and wave height and direction; need to be measured along the objects trajectory for a wide range of conditions.

@@

where $V$ is the velocity of the drifting object, $V_c$ is surface current velocity, $V_l$ is the 'leeway' velocity and $V_w$ is the velocity due to wave forces. As stated, $V_c$ is the velocity of the surface current and is averaged over the vertical extend of the drifting object. Velocity of the surface current is expressed as $$V_c = V_5 + V_s$$
where $V_5$ is the average velocity for the the top 5m of the ocean (estimated by an ocean model) and $V_s$ is the velocity effect of Stokes drift due to wave action. It is important to note that the approximation of Stokes drift does not take into account finer-scale vertical differences of the velocity field, such as Coriolis-Stokes forcing. Stokes drift generally manifests in hydrodynamic environments with short-wavelengths and locally-generated waves. @griffin2017 mitigated the exclusion of finer-scale drivers of Stokes Drift by comparing modeled surface velocities with those of *in situ* drifters.

$V_l$ is the velocity of the object relative to the water due to effects of wind force directly on the object and by expressed by $$V_l = w U_10$$

# Particle dispersion modelling

## Particle dispersion models

A study by @lebreton2012 proposed a methodology to track floating debris from source to sink based on descriptions of global waste production and ocean surface currents, using a combination of hydrodynamic and particle dispersion modelling. The particle tracking model applied uses a two stage process, (1) a hydrodynamic model solves the equations of motion to characterise the movement of water within the domain of the model, (2) virtual particles are released into the flow field established in the first step, and particles are allowed to move through hydrodynamic forcing. The authors extracted sea surface currents from the  HYCOM/NCODA ocean circulation modelling system. The HYCOM model is forced by the US Navy's Operational Global Atmospheric Prediction System (NOGAPS). The model also takes into account wind stress, wind speed, heat flux, and precipitation; and was looped five times to accurately represent ocean circulation patterns for 30 years. The authors used the velocity data aquired from HYCOM and coupled it to the Lagrangian particle tracking model Pol3DD which was used to simulate dispersion. No additional wind stress terms were applied to the motion of particles, as this factor was already expressed in the HYCOM hydrodynamic data. Potential extra stress for emerged parts of debris was considered negligible as all floating debris are assumed to be fully submerged. The Pol3DD model was also modified so provide additional information on individual particles such as origin, age, and trajectory information of individual particles. Since the model considered a spherical earth, particles were allowed to re-enter the simulation if the passed around the 'globe'. In this study only the surface layer was considered as the principle layer which determines the fate of floating particles i.e anthorpogenic debris.

@griffin2017 added to the search for flight MH370 by providing estimates of the trajectory of the debris. This was achieved through a combination of an advanced ocean model calibrated with oceanic drifters, and drifters in the form of aircraft parts.

Dispersal is recognised as important driver of biodiversity, composition and structure of ecological systems in the marine environment [@bernardes2018;@helmuth1994;@highsmith1985;macarthur2001;@jackson2010]. These processes are largely dependent on currents for dispersal of migrant populations which ultimately promote connectivity of marine ecosystems [@bernardes2018;@mackas1985;@mackas1985;@zakas2009]. This is particularly true for organisms which lack pelagic larvae, and are reliant on other modes of transport to reach new habitats through passive modes of dispersal from anthropogenic and natural sources. Anthropogenic sources include marine litter and structural debris while examples of organic sources include pumice and macroalgae in the form of kelp-rafts. Kelp-rafts have been identified as an important mode of passive dispersal of marine organisms, such as invertebrates [@helmuth1994; whichmann2012], epiphytes [@], grazers [@nikula2010], invasive species [@lewis2005] and macroalgae themselves [@macaya2005] which either make landfall or are transported offshore where they eventually sink. A kelp-raft consists of an entanglement of macroalgae (one or multiple species) which have been dislodged from the benthic environment through hydrodynamic forces (mostly storms) and are positively buoyant by means of gas-filled pneumatocysts and/or reproductive organs. Kelp-rafts are capable of travelling vast distances [@] and are considered important dispersal vectors in temperate latitudes, such as the Southern California Bight [@hobday2000] and the Northern Baltic Sea [@].

The trajectory of kelp rafts are determined by a combination of oceanographic conditions and biological processes. Since kelp rafts float, evidence points largely to wind driven surface currents as the main driver, although the relative importance of wind versus surface currents is largely unknown. For example a study by @harrold1989 used radio-trackers on both natural and artificial *Macrocystis pyrifera* kelp rafts to investigate seasonal trajectory patterns. The results showed that tracked kelp rafts were mostly deposited nearby the source population and that seasonal wind direction was the primary driver of trajectory for kelp rafts submerged roughly ~0.5 - 1m below the surface @harrold1989. However, the authors did note the influence of surface currents for some of their experiments when winds relaxed; which suggests that surface currents may play in determining trajectory for kelp rafts less exposed to wind or when prevailing winds are low.

Kelp-raft abundance has been shown to vary temporally and spatially in the ocean and around coastlines. The temporal variability is dependent on seasonal growth patterns. For example, studies investigating the dispersal patterns of *Sargassum* in the West Pacific show increased abundance during growth seasons (spring and summer) as more biomass is available to fragment [@deysher1981; @kingsford1992]. Temporal variability may also be related to seasonal storm frequency where more storms in a particular season relate to an increase in kelp-raft abundance. For example a study by @kingsford1995 investigated the contribution of *Macrocystis pyrifera* rafts to habitat complexity in pelagic environments. The authors found that the abundance of *M. pyrifera* rafts increased during seasons where storms occurred more frequently.

>>>>

---
title: "Quantifying passive kelp rafting phenomena and connectivity of nearshore communities around the coast of South Africa:Development of a 3D spectral numerical coastal-flow model"
author: 
- RM Coppin^[University of the Western Cape]
- C Rautenbach^[South African Weather Service]
- AJ Smit^[University of the Western Cape]
output:
  html_document:
    df_print: paged
    toc: yes
    theme:  simplex
  word_document:
    fig_caption: yes
    reference_docx: chapter_1_template.docx
    toc: yes
  pdf_document:
    toc: no
bibliography: references.bib
---

@@

Marine ecosystems are maintained by a variety of complex interactions between abiotic and biotic variables such as temperature, wave exposure, pH, competition, and processes such as top-down and bottom-up control, predator-prey relationships and phenology [@doney2011; @Harley2012; @Burrows2011; @McGowan1998]. These abiotic and biotic variables, the interactions between them, and the various ecological processes, ultimately determine the community composition and ecological functioning of all ecosystems [@Guimaraes1996; @Harley2006; @Poloczanska2013; @Jennings2010; @Polovina2005; @Johnson2011; @Krumhansl2016; @Dayton1999]. Climate directly and indirectly affects the way in which abiotic and biotic variables interact, but is often compounded by other impacts such as habitat destruction, pollution, and over-fishing [@Blamey2012; @Blamey2015]. Temperature and wave exposure have been recognised as important variables with regards to climate-driven changes within the ocean (Guimaraes & Coutinho 1996; McGowan et al. 1998; McQuaid & Branch 1984; Laufkötter et al. 2015; Belkin 2009; Filbee-Dexter et al. 2018; Miller et al. 2011; Smale et al. 2011; Smale & Moore 2017). In order to persist and survive within variable and changing environments, organisms must either migrate, adapt, or die.

Seaweeds are sessile organisms that are unable to migrate to new areas when local environmental conditions become unsuitable and therefore are forced to adapt to new conditions in order to avoid expiration (Dayton 1985; Tegner et al. 1997). The main form of mortality for seaweeds is through mechanical dislodgement by wave action (Schiel et al. 2006; Seymour et al. 1989; Thomsen et al. 2004; Cavanaugh et al. 2011; Graham et al. 1997; Demes et al. 2013; Edwards & Estes 2006). Seaweeds, particularly brown seaweeds, are able to undergo rapid morphological adaptation to the hydrodynamic environment (Miller et al. 2011; Friedland & Denny 1995; Wing et al. 2007; Wernberg & Thomsen 2005; Fowler-Walker et al. 2006; Dudgeon & Johnson 1992; Blanchette et al. 2002). This allows seaweeds to reduce mortality through mechanical dislodgement by inducing morphology which reduces overall drag. Seaweeds that are unable to avoid mechanical dislodgement either raft out to sea or wash up onto beaches. Not all beach-cast kelp may have originated from a nearby kelp population and may have originated from other sites or regions of the coast through rafting and ocean currents (Malm et al. 2004; Emery & Tschudy 1941; Filbee-Dexter & Scheibling 2012; Hobday 2000). Therefore, because kelp morphology is specific to its local environment the morphological features may be able to indicate, within a certain amount of probability, what site or region it most likely originated from. In other words, beach-cast kelp may be used as proxy for investigating the flow of coastal currents. Using kelp as a proxy for determining its original location will be calibrated by means of a hydrodynamic model which will be designed from already existing SWAN and Delft3D models. This combined approach will allow investigation into flow regimes around the west and south-west coasts of South Africa and the role they play in subtidal ecology. Furthermore, the model could be used to determine transport of plastic pollution around the coast. For instance, micro-plastics are recognised as a threat to marine life (Seltenrich 2015; Azzarello & Van Vleet 1987; Wright et al. 2013; Derraik 2002; Andrady 2011; Fendall & Sewell 2009; Ivar Do Sul & Costa 2014; O ’donoghue & Marshall n.d.; Ballent et al. 2012), however very little is known about how microplastics may be transported along the coast.

@@

In the wave swept nearshore one would expect organisms to reflect the harsh hydrodynamic environment by being streamlined, small and amoured. This is certainly the case for an array of fauna which often comprise of hard, rigid bodies that are held firmly in place to the substratum such as limpets and isopods......

@@

```{r surface elevation figures, echo=FALSE, warning=FALSE, fig.align="center", fig.cap="\\label{fig:surface elevation figures}The definition of a wave in a time record of the surface elevation with downward zero-crossings (upper panel) and upward zero-crossings (lower panel)."}

@@

```{r Hs_Tp_surface_elevation figure, echo=FALSE, warning=FALSE, fig.align="center", fig.cap="The definition of wave height and wave period in a time record of the surface elevation (the wave is defined with downwrd zero-crossings."}

@@

# Ocean modelling

An ocean model is essentially a representation of physical processes in the ocean in the form of equations or computer code which aids in furthering our understanding of how the ocean works. Examples of physical processes include the exchange of energy, mass, and momentum between the ocean and external drivers; ocean movement/dynamics; and 3D mixing and dissipation processes. Examples of process which exchange energy, mass, and momentum between sources include, but are not limited to, radiation, evaporation, precipitation, river runoff and wind energy. Ocean movement and dynamics, as the name suggests, refers to processes which determine the movement of water in the ocean, such as horizontal advection and vertical convection. The 3D mixing and dissipation refers to processes that remove energy from the system such as turbulence caused by temperature differences and wind forcing. There are several approaches to modelling the ocean all of which have advantages and limitations. Conceptual or process models are limited in the different processes that can be incorporated into the model and are often regarded as simple, low resolution models. However conceptual or process models make it easier to interpret dynamics and processes in isolation, and can be run parallel to other models, and can accommodate larger simulations compared to more complex models. In essence, conceptual or process models are mathematical tools to aid in contributing to the theory of ocean processes. Another type of approach are Earth System Models of Intermediate Complexity (EMICs), which have a higher resolution and can run for long timescales with relatively low computation cost. These approaches have been used to investigate coupling of climate and abiotic processes, such as ice sheets and carbon feedback loops. Although EMICs incorporate more complexity, these approaches do not accurately reflect natural ocean processes; instead General Circulation Models (GCMs) provide a better description of the ocean. General circulation models are capable of incorporating multiple processes at a higher resolution. Generally speaking, currently GCMs provide the most accurate way of describing natural ocean processes.

Ocean models are popular tools for investigating future climate scenarios and has been used mostly in this regard. However, ocean models can also be used in operational oceanography (now-casts and forecasts) in the maritime and shipping industry, experimental oceanography, and can be used to mechanistically interpret ocean observations. Ocean models are similar to atmospheric models but fundamentally differ in some respects. For example, atmospheric models involve air which is a compressible gas while ocean models involve seawater which is nearly in-compressible. These differences have significant impacts for how aspects of volume, temperature and pressure influence one another. In air, the interplay between these factors is largely linear and well understood, while in the ocean the relationship is non-linear and poorly understood. For ocean models the equation state used is $$\rho = f(x)TSP$$
where, $\rho$ represents density and is a function of, $T$ temperature, $S$ salinity, and $P$ pressure. Although the equation state is more complex compared to that of air some aspects are slightly simpler, as seawater is in-compressible the water entering a grid box will be the same as the water exiting the grid box. Another major difference between ocean and atmospheric models is in ocean models salinity must be taken into account while with atmospheric models humidity is an important factor. Ocean and atmospheric models also differ slighlty in terms of vertical structure i.e. layering. The surface of the ocean is where most heating and coolong occurs which cause the formation of a marine boundry or mixed layer. Below the mixed layer the ocean becomes more stratified which is ideal for determining flow. However, complications arise in the horizonal layer which does not continiously layer like the atmosphere. The horizontal layer in ocean models needs to take into account irregular bathymetry, shape and size of continental shelves/basins. Furthermore, run-off from rivers can cause changes in the density of seawater along the coasts, thereby altering their flow regimes. Therefore, ocean models need to resolve both the vertical and horizontal structuring, particulary along the ocean margins, which can be achieved through selecting the correct horizontal and vertical co-ordinate systems. Finally, the ocean interior is largely driven by density gradients compared to the atmosphere which is well mixed and so the equalibrium timescales are much slower relative to the atmosphere. For instance, an ocean model can take as long as 1000 years to start an ocean model from rest.

Ocean models use similar equations to atmospheric models, with some minor differences (might add equation list). Theses equations form the basis of a ocean model, but more input is needed for a model to run successfully. Ocean models also need input on boundary conditions such as basin geometry, bathymetry and atmospheric pressure. The Atmospheric pressure can be solved for dynamically, or at the very least, the effects are approximated and imposed. Initial abiotic conditions are another important input into ocean models. Initial conditions include the mean state of temperature, salinity and velocity fields and can be calculated from climatology data or from a previously run ocean model. Finally, the forcing fields, both dynamic and static, are needed such as shortwave radiation, long-wave radiation, latent heat, evaporation, precipitation and land surface run-off. Examples of dynamic forcing fields include winds and tides.

A model can be defined by the resolution, parametrisation and intergration.

## Introduction
# Hydrodynamic modelling
## Introduction

@@

# Particle dispersion modelling
## Particle dispersion models

A study by @lebreton2012 proposed a methodology to track floating debris from source to sink based on descriptions of global waste production and ocean surface currents, using a combination of hydrodynamic and particle dispersion modelling. The particle tracking model applied uses a two stage process, (1) a hydrodynamic model solves the equations of motion to characterise the movement of water within the domain of the model, (2) virtual particles are released into the flow field established in the first step, and particles are allowed to move through hydrodynamic forcing. The authors extracted sea surface currents from the  HYCOM/NCODA ocean circulation modelling system, which is forced by the US Navy's Operational Global Atmospheric Prediction System (NOGAPS). The model also took into account wind stress, wind speed, heat flux, and precipitation and was looped five times to accurately represent ocean circulation patterns for a 30 year period. The authors used the velocity data aquired from HYCOM and coupled it to the Lagrangian particle tracking model Pol3DD which was used to simulate dispersion. No additional wind stress terms were applied to the motion of particles, as this factor had already been incorporated in the HYCOM hydrodynamic data. Potential extra stress for emerged parts of debris was considered negligible as all floating debris are assumed to be fully submerged. The Pol3DD model was also modified so provide additional information on individual particles such as origin, age, and trajectory information of individual particles. Since the model considered a spherical earth, particles were allowed to re-enter the simulation if they passed around the 'globe'. Important to note is only the surface layer was modelled in this study, as much of the evidence identifies surface currents as the main dispersal vector of anthropogenic debris.

@griffin2017 added to the search for flight MH370 by providing estimates of the trajectory of the debris. This was achieved through a combination of an advanced ocean model calibrated with oceanic drifters, and drifters in the form of aircraft parts. Unlike previous work regarding the recovery of MH370 wreckage, this study incorporated windage factors into the trajectory model as well as providing better estimates of Stokes Drift.

Dispersal is recognised as important driver of biodiversity, composition and structure of ecological systems in the marine environment [@bernardes2018;@helmuth1994;@highsmith1985;@macarthur2001;@jackson2010]. These processes are largely dependent on currents for dispersal of migrant populations which ultimately promote connectivity of marine ecosystems [@bernardes2018;@mackas1985;@mackas1985;@zakas2009]. This is particularly true for organisms which lack pelagic larvae, and are reliant on other modes of transport to reach new habitats through passive modes of dispersal from anthropogenic and natural sources. Anthropogenic sources include marine litter and structural debris while examples of organic sources include pumice and macroalgae in the form of kelp-rafts. Kelp-rafts have been identified as an important mode of passive dispersal of marine organisms, such as invertebrates [@helmuth1994; whichmann2012], epiphytes [@], grazers [@nikula2010], invasive species [@lewis2005] and macroalgae themselves [@macaya2005] which either make landfall or are transported offshore where they eventually sink. A kelp-raft consists of an entanglement of macroalgae (one or multiple species) which have been dislodged from the benthic environment through hydrodynamic forces (mostly storms) and are positively buoyant by means of gas-filled pneumatocysts and/or reproductive organs. Kelp-rafts are capable of travelling vast distances [@] and are considered important dispersal vectors in temperate latitudes, such as the Southern California Bight [@hobday2000] and the Northern Baltic Sea [@].

Kelp-raft abundance has been shown to vary temporally and spatially in the ocean and around coastlines, which is dependent on seasonal growth patterns. For example, studies investigating the dispersal patterns of *Sargassum* in the West Pacific show increased abundance during growth seasons (spring and summer) as more biomass is available to fragment [@deysher1981; @kingsford1992]. Temporal variability may also be related to seasonal storm frequency where more storms in a particular season relate to an increase in kelp-raft abundance. For example a study by @kingsford1995 investigated the contribution of *Macrocystis pyrifera* rafts to habitat complexity in pelagic environments. The authors found that the abundance of *M. pyrifera* rafts increased during seasons where storms occurred more frequently.

The trajectory of kelp rafts are determined by a combination of oceanographic conditions and biological processes. Since kelp rafts float, evidence points largely to wind driven surface currents as the main driver, although the relative importance of wind versus surface currents is largely unknown. For example a study by @harrold1989 used radio-trackers on both natural and artificial *Macrocystis pyrifera* kelp rafts to investigate seasonal trajectory patterns. The results showed that tracked kelp rafts were mostly deposited nearby the source population and that seasonal wind direction was the primary driver of trajectory for kelp rafts submerged roughly ~0.5 - 1m below the surface @harrold1989. However, the authors did note the influence of surface currents for some of their experiments when winds relaxed; which suggests that surface currents may play a larger role in determining trajectory for kelp rafts less exposed to wind or when prevailing winds are low.

# Calculating characteristics of kelp drift
## Buoyancy of drifting objects in the oceans

@@

```{r BF_1, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 3, fig.height = 3, fig.align="center", fig.cap='\\label{fig:BF_1}Conceptual model of bouyancy characteristics.'}

@@

The effects that wind produce on floating objects is sometimes referred to as "windage factors", and is difficult to accurately and empirically describe. The difficulty is due to accurate measurements needed for the current velocity, wind velocity, wave height and wave direction. Furthermore, these factors do not act independently and instead influence one another, which compounds the complexity of calculating wind effects on drifting objects. To overcome this challenge, measurements of current and wind velocity, and wave height and direction are measured along the objects trajectory for a wide range of conditions.

@@

where $V$ is the velocity of the drifting object, $V_c$ is surface current velocity, $V_l$ is the 'leeway' velocity and $V_w$ is the velocity due to wave forces. As stated, $V_c$ is the velocity of the surface current and is averaged over the vertical extend of the drifting object. Velocity of the surface current is expressed as $$V_c = V_5 + V_s$$
where $V_5$ is the average velocity for the the top 5m of the ocean (estimated by an ocean model) and $V_s$ is the velocity effect of Stokes Drift due to wave action. It is important to note that the approximation of Stokes drift does not take into account finer-scale vertical differences of the velocity field, such as Coriolis-Stokes forcing. Stokes drift generally manifests in hydrodynamic environments with short-wavelengths and locally-generated waves. @griffin2017 mitigated the exclusion of finer-scale drivers of Stokes Drift by comparing modeled surface velocities with those of *in situ* drifters.

$V_l$ is the velocity of the object relative to the water due to effects of wind force directly on the object and can expressed by $$V_l = w U_10$$
where $w$ is a linear 'windage' or 'leeway' coefficient and $U_10$ is the wind speed at 10m height. The study by @griffin2017 used a 10m height to model drifing airplane debris, however this may be adjusted according to drifting object being modelled. The nature of $w$ (linear windage) is dependent on the object, for instance yachts tend to drift in a particular orientation to the wind, and so the angle of the drifting object is dependent on said orientation. Furthermore, drifting object which are more wind exposed and are less submerged will have high windage. Due to these factors, and to allow for a non-zero angle between $V_l$ and $U_10$, $w$ (linear windage) can be a complex number.

$V_W$ is the velocity of the object as a result of direct wave force, and is dependent on the size and shape of the object. In instances when wave-wave interactions result in an small, unbalanced force then $V_W$ is assumed to be zero. Also, objects may be able to absorb wave energy which can result in a non-zero $V_W$ value. Since surface-waves are wind driven, the $V_W$ is calculated by $$V_W = f(U_10)$$

### 2019-11-08 morphology_paper.Rmd
added section to lit review, updated outputs, and added references to bib. file

---
title: Using numerical modelling to investigate temperature and wave exposure as drivers
  of macroalgal morphological characteristics
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
abstract: "**Abstract**: Macroalgae morphological characteristics are determined by a combination of environmental factors with wave exposure and temperature being identified as the main drivers of morphological adaptation. Macroalgae have shown to exhibit different responses to different magnitudes of exposure, such as reduction in overall size and strength increasing traits. In terms of temperature, warmer environments have been shown to reduce the overall size of resident and transplanted species. However, none of the past studies have identified specific wave and temperature metrics responsible for the morphological adaptation macroalgae exhibit. Past research has often used simple or two dimensional models of wave exposure which does not take into account important aspects of the nearshore environment such as wave breaking, refraction and diffraction. Furthermore, past studies have often used satellite as sources for temperature data which has been shown to have large bias when measuring the nearshore environment. This study used *in situ* temperature data and wave power calculated from a 3D-numerical model to identify specific temperature and wave metrics responsible for morphological adaptation.  The results identify wave exposure as the main driver of morphological adaptation while identifying specific wave metrics. Furthermore, the results show differences in wave metrics between species; and between deep and shallow populations."
fontsize: 10pt
output:
  word_document:
    reference_docx: frontiers_template.docx
  pdf_document: 
    fig_caption: yes
geometry: margin=1in
papersize: A4
tables: yes
bibliography: references.bib
---

>>>>

---
title: Using numerical modelling to investigate temperature and wave exposure as drivers
  of macroalgal morphological characteristics
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
bibliography: references.bib
output:
  html_document:
    df_print: paged
  pdf_document:
    fig_caption: yes
  word_document:
    reference_docx: frontiers_template.docx
geometry: margin=1in
fontsize: 10pt
papersize: A4
tables: yes
abstract: 'Macroalgae morphological characteristics are determined by
  a combination of environmental factors with wave exposure and temperature being
  identified as the main drivers of morphological adaptation. Macroalgae have shown
  to exhibit different responses to different magnitudes of exposure, such as reduction
  in overall size and strength increasing traits. In terms of temperature, warmer
  environments have been shown to reduce the overall size of resident and transplanted
  species. However, none of the past studies have identified specific wave and temperature
  metrics responsible for the morphological adaptation macroalgae exhibit. Past research
  has often used simple or two dimensional models of wave exposure which does not
  take into account important aspects of the nearshore environment such as wave breaking,
  refraction and diffraction. Furthermore, past studies have often used satellite
  as sources for temperature data which has been shown to have large bias when measuring
  the nearshore environment. This study used *in situ* temperature data and wave power
  calculated from a 3D-numerical model to identify specific temperature and wave metrics
  responsible for morphological adaptation.  The results identify wave exposure as
  the main driver of morphological adaptation while identifying specific wave metrics.
  Furthermore, the results show differences in wave metrics between species; and between
  deep and shallow populations.'
---
